---
title: "plot genotype data for multiple models"
author: "Andrea Chen"
date: "`r Sys.Date()`"
output: pdf_document
---

# parameters for plotting
```{r setup, include=FALSE}
wd_path <- normalizePath(dirname(rstudioapi::getActiveDocumentContext()$path))
setwd(wd_path)
base_plot_folder <- "../plots/1_23/constant_env"
if (!dir.exists(base_plot_folder)) {
  dir.create(base_plot_folder, recursive = TRUE)
}
hcl_color_palette <- "Dark 2" # color pattern for lines 
```

# load genotype data
```{r}
geno_df <- readRDS("../r_data/genotype_data.rds")
pheno_df <- readRDS("../r_data/phenotype_data.rds")
```

```{r}
geno_df <- genotype_df
unique(geno_df$psi)
```
# parameter space for data
```{r}
shift_size <- "no_sz"
for (gen in c(9999)){
  for (n in c(2,8,20,50)){
    # for (shift_size in c(40, 80, 120)){
      for (reg in c("fit", "ext", "ave")){
        for (type in c("genotype", "phenotype")){
          cat("Plotting:", type, " generation:", gen, " n:", n, " shift_size:", shift_size, " reg:", reg, "\n")
          
          breaks_vec <- seq(-50, 50, by = 1)  
          if (type == "genotype"){
            df_to_use <- geno_df
          } else if (type == "phenotype"){
            df_to_use <- pheno_df
          }

          if(gen == 10000){
            title <- paste0(type, "_", gen, "/", reg, "_", shift_size,"_", n, "opt = ", shift_size)
          }else{
            title <- paste0(type, "_", gen, "/", reg, "_", shift_size,"_", n, "opt = ", 0)
          }

          # filt_A <- list(reg=reg, n=n, psi=0, shift_size=shift_size)
          # filt_B <- list(reg=reg, n=n, psi=0.5, shift_size=shift_size)
          # filt_C <- list(reg=reg, n=n, psi=0.9, shift_size=shift_size)

          filt_A <- list(reg=reg, n=n, psi=0)
          filt_B <- list(reg=reg, n=n, psi=0.5)
          filt_C <- list(reg=reg, n=n, psi=0.9)

          png(file.path(base_plot_folder, paste0(type, "_", gen, "/", reg, "_", shift_size,"_", n, ".png")), width=2200, height=1500, res=150)
          if (!dir.exists(file.path(base_plot_folder, paste0(type, "_", gen)))) {
            dir.create(file.path(base_plot_folder, paste0(type, "_", gen)), recursive = TRUE)
          }
          par(cex.axis = 1.6, cex.lab = 2.0, cex.main = 1.0, mar = c(4.5, 6, 3, 1) + 0.1)
          fils <- list(filt_A, filt_C)
          cols <- hcl.colors(3, hcl_color_palette, rev = TRUE)
          cols <- c(cols[3], cols[1])
          plot_hist_lines_multi(
            df = df_to_use,
            # filters = list(filt_A, filt_B, filt_C),
            filters = fils,
            generation = gen,
            breaks = breaks_vec,
            labels = list(title = title,
                          x = ifelse(type == "genotype", "Genotype value", "Phenotype value"),
                          y = "Proportion"),
            show_rep = FALSE,
            ymax = if (type == "phenotype") 0.25 else 0.05,
            cols = cols
          )

          leg_expr <- as.expression(lapply(lapply(fils, function(f) f$psi), function(v) bquote(psi == .(v))))
                
          legend("topright", bty = "n",
                  legend = leg_expr,
                col = cols, lwd = 2, cex=2.2)
          dev.off()
        }
      }
    }
  }
# }
```

```{r}
unique(geno_df[geno_df$reg == "inv", ]$generation)
# colnames(geno_df)
```

## plot multi generation 
```{r}
breaks_vec <- seq(-40, 150, by = 1)                 # outside [-5,5] ignored
filt <- list(reg="inv", n=8, shift_size=120, beta=0.0)
gens <- c(9999, 10010, 10020, 10030, 10050, 10070, 10100)

col = grDevices::hcl.colors(3, "Dark 2", rev = TRUE)[3]

out_dir  <- file.path(base_plot_folder, "geno_multi")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
out_file <- file.path(out_dir, paste0( reg, "_geno.png"))

png(out_file, width=2200, height=1500, res=150)
par(cex.axis = 1.6, cex.lab = 2.0, cex.main = 1.0, mar = c(4.5, 6, 3, 1) + 0.1)
plot_multigen_binned_single_filter(
  df = geno_df,
  filter = filt,
  generations = gens,
  breaks = breaks_vec,
  title = bquote("Binned distributions (" * psi==.(filt$beta) * ")"),
  xlab  = "Genotype value",
  ylab  = "Proportion",
  col   = col,
  show_rep = FALSE,
  legend_pos = "topright",
  legend_as_offset = TRUE,
  ymax = 0.2
)
leg_expr <- as.expression( bquote(psi == .(filt$beta)))
                
# legend("topright", bty = "n",
#         legend = leg_expr,
#        col = grDevices::hcl.colors(2, hcl_color_palette, rev = TRUE), lwd = 2, cex=2.2)
dev.off()
cat("Saved:", out_file, "\n")
```


breaks_vec <- seq(-5, 5, by = 0.1)                 # outside [-5,5] ignored
filt <- list(reg="inv", n=8, shift_size=120, beta=0.0)
gens <- c(9999, 10050, 10100)

out_dir  <- file.path(base_plot_folder, "geno_multi")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
out_file <- file.path(out_dir, paste0("geno_multi/", reg, "_geno.png"))

png(out_file, width=2200, height=1500, res=150)
par(cex.axis = 1.6, cex.lab = 2.0, cex.main = 1.0, mar = c(4.5, 6, 3, 1) + 0.1)
plot_multigen_binned_single_filter(
  df = geno_df,
  filter = filt,
  generations = gens,
  breaks = breaks_vec,
  title = bquote("Binned distributions (" * psi==.(filt$spsi) * ")"),
  xlab  = "Genotype value",
  ylab  = "Proportion",
  col   = "steelblue",
  show_rep = TRUE
)
leg_expr <- as.expression( bquote(psi == .(filt$beta)))
                
legend("topright", bty = "n",
        legend = leg_expr,
       col = grDevices::hcl.colors(2, hcl_color_palette, rev = TRUE), lwd = 2, cex=2.2)
dev.off()
cat("Saved:", out_file, "\n")










