---
title: "plot genotype data for multiple models"
author: "Andrea Chen"
date: "`r Sys.Date()`"
output: pdf_document
---

# parameters for plotting
```{r setup, include=FALSE}
wd_path <- normalizePath(dirname(rstudioapi::getActiveDocumentContext()$path))
setwd(wd_path)
base_plot_folder <- "../plots/evolve_sociality"
hcl_color_palette <- "Dark 2" # color pattern for lines 
```

# load genotype data
```{r}
geno_df <- readRDS("genotype_data.rds")
```

# parameter space for data
## all psi values to plot for fittest model
psi <- c(0.0, 0.1, 0.3, 0.5, 0.7, 0.9, 1.0) 
## all beta values to plot for inverse-variance model
beta <- c(0.0, 0.01, 0.03, 0.05, 0.07, 0.3, 0.5, 0.7, 0.9, 1.0)
## speficy the model to plot; "inv" = inverse variance model, "fit" = fittest individual model
regimes <- c("inv", "fit")
## all group sizes
plot_n <- c(2, 8, 20, 50)
## all shift sizes 
plot_shift_sizes <- c(40,80,120) 
## all generations
gen_seq = c(seq(9999, 10010, by = 1), seq(10010, 10100, by=10)) # generations to plot

# change parameters here for plotting a subset of the data
## PARAMETER: single panel multi generation plot
```{r}
param_val <- c(0.0, 0.5)
plot_n <- c(50)
plot_shift_sizes <- c(120)
regimes <- c("fit")
num_bins <- 140
gen_seq = seq(10010, 10030, by = 10)
include_pre_selection <- TRUE
single_panel <- TRUE
```

# PARAMETER: multi panel multi generation plot
```{r setup, include=FALSE}
param_val <- c(0.0) # all beta values to plot for inverse-variance model
plot_n <- c(2, 8, 20, 50) # group sizes to plot for both models
plot_shift_sizes <- c(40,80,120, 300) # shift sizes to plot for both models
regimes <- c("fit", "ave", "ext", "inv")  # "inv" = inverse variance model, "fit" = fittest individual model
single_panel <- TRUE  
include_pre_selection <- FALSE
num_bins <- 120
gen_seq = seq(10000, 10030, by = 10)
```

# RUN PLOTTING after change the parameters above
```{r}
min_bin <- -as.integer(num_bins/2)
max_bin <- as.integer(num_bins/2)
bin_breaks <- seq(min_bin, max_bin, length.out = num_bins + 1)

# Derive a constant bin width from your reference breaks
bin_width <- (max(bin_breaks) - min(bin_breaks)) / num_bins
half_span <- (num_bins * bin_width) / 2

if (regimes != "inv") {
  plot_psi <- param_val
} else {
  plot_beta <- param_val
}

geno_df <- genotype_df

# ---- Run ----
plot_genotype_multi_by_beta(
    geno_df,
    out_subfolder = paste0("genotype_overtime_", paste0(c(min(gen_seq), max(gen_seq)), collapse = ":")),
    gen_seq = gen_seq,
    num_bins = num_bins,
    bin_width = bin_width,
    half_span = half_span,
    regimes = regimes,
    single_panel = single_panel,
    include_pre_selection = include_pre_selection
)
```
