---
title: "plot genotype, phenotype and effect size data"
author: "Andrea Chen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
setwd("/workdir/zc524/collective_behavior_evo/r_remote")
base_data_folder <- "../data/n_group_dist_mean(mislabel_reg)"
base_plot_folder <- "../r_plots/n_group_dist_mean/before_shift"
target_psi <- c(0.0, 0.3, 0.5, 0.7, 0.9)
regimes <- c("fit")  # "ave" = average, "fit" = fitness-based
target_gen <- 9999
```

# load GENOTYPE data
```{r}
# Set the top-level folder
folder <- paste0(base_data_folder, "/genotype")

# List all .tsv files recursively
file_list <- list.files(
  path = folder,
  pattern = "\\.tsv$",
  full.names = TRUE,
  recursive = TRUE
)

# Function to extract metadata from path
# Matches: genotype/n_<n>_psi_<psi>_sz_<sz>_reg_<ave|fit>_<rep>.tsv
extract_geno_metadata <- function(path) {
  pat <- "genotype/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)\\.tsv"
  matches <- regexec(pat, path)
  parts <- regmatches(path, matches)[[1]]

  if (length(parts) != 6) {
    warning("Could not extract metadata from path: ", path,
            " with matches: ", paste(parts, collapse = ", "))
    return(data.frame())
  }

  data.frame(
    n    = as.integer(parts[2]),
    psi  = as.numeric(parts[3]),
    sz   = as.integer(parts[4]),
    reg  = parts[5],             # "ave" or "fit"
    rep  = as.integer(parts[6]),
    stringsAsFactors = FALSE
  )
}

# Helper: get shift_size from the companion "data" file (optimum at generation == 10001)
get_shift_from_data <- function(geno_file, base_data_folder, opt_gen = 10001L) {
  # Mirror the path: .../genotype/... -> .../data/...
  data_path <- sub("/genotype/", "/data/", geno_file, fixed = TRUE)
  if (!file.exists(data_path)) {
    data_path <- file.path(base_data_folder, "data", basename(geno_file))
  }

  if (!file.exists(data_path)) {
    warning("Data file not found for ", basename(geno_file), " (looked for: ", data_path, ")")
    return(NA_real_)
  }

  d <- tryCatch(read.table(data_path, header = TRUE, sep = "\t"), error = function(e) NULL)
  if (is.null(d)) {
    warning("Failed to read data file: ", data_path)
    return(NA_real_)
  }
  if (!("tick" %in% names(d)) || !("optimum" %in% names(d))) {
    warning("Missing 'tick' or 'optimum' column in data file: ", data_path)
    return(NA_real_)
  }

  row10001 <- d[d$tick == opt_gen, , drop = FALSE]
  if (nrow(row10001) < 1) {
    warning("No row with tick == ", opt_gen, " in ", data_path)
    return(NA_real_)
  }

  val <- suppressWarnings(as.numeric(row10001$optimum[1]))
  if (!is.finite(val)) {
    warning("Non-numeric 'optimum' at generation ", opt_gen, " in ", data_path)
    return(NA_real_)
  }
  val
}

# Read and combine all files
opt_gen <- 10001L
genotype_df <- do.call(rbind, lapply(file_list, function(file) {
  df <- tryCatch(read.table(file, header = TRUE, sep = "\t"), error = function(e) NULL)
  if (is.null(df)) {
    warning("Failed to read genotype file: ", file)
    return(NULL)
  }

  meta <- extract_geno_metadata(file)
  if (nrow(meta) == 0) return(NULL)

  # Attach metadata
  df$rep        <- meta$rep
  df$psi        <- meta$psi
  df$n          <- meta$n
  df$reg        <- meta$reg        # <- new column ("ave" or "fit")
  df$sz_orig    <- meta$sz         # preserve original sz
  df$shift_size <- if (meta$sz == 0L) get_shift_from_data(file, base_data_folder, opt_gen) else meta$sz
  df$file_id    <- basename(file)
  df
}))

```

# plot GENOTYPE histogram
```{r}
library(RColorBrewer)
# Assuming you want to focus on a specific generation (e.g., 10100)
df <- subset(genotype_df, generation == target_gen)

# Define histogram bins
num_bins <- 120
bin_breaks <- seq(-60, 60, length.out = num_bins + 1)
bin_mids <- 0.5 * (head(bin_breaks, -1) + tail(bin_breaks, -1))

# Color setup
psi_vals <- target_psi
colors <- rainbow(length(psi_vals))
names(colors) <- as.character(psi_vals)
```

## plot average GENOTYPE for each psi
```{r}
# Assumes genotype_df already has: generation, n, psi, rep, shift_size, sz_orig, reg, X* columns
# Also assumes you have: target_gen, num_bins, bin_breaks, bin_mids, base_plot_folder, regimes

avg_curve_for_subset <- function(df_block) {
  # df_block is already filtered to a single (panel × psi × reg) subset at target_gen
  if (nrow(df_block) == 0) return(rep(NA_real_, length(bin_mids)))
  counts_mat <- matrix(NA_real_, nrow = nrow(df_block), ncol = length(bin_mids))
  x_cols <- grep("^X", names(df_block))
  for (i in seq_len(nrow(df_block))) {
    vals <- suppressWarnings(as.numeric(as.vector(unlist(df_block[i, x_cols]))))
    vals <- vals[is.finite(vals)]
    if (length(vals) == 0) next
    h <- hist(vals, breaks = bin_breaks, plot = FALSE)
    if (sum(h$counts) == 0) next
    counts_mat[i, ] <- h$counts / sum(h$counts)
  }
  colMeans(counts_mat, na.rm = TRUE)
}

for (reg_lab in regimes) {
  df_reg_all <- subset(genotype_df, generation == target_gen & reg == reg_lab)
  if (nrow(df_reg_all) == 0) next

  for (n_val in sort(unique(df_reg_all$n))) {
    df_n <- subset(df_reg_all, n == n_val)
    if (nrow(df_n) == 0) next

    # Panels: one pooled panel for sz_orig==0, plus one per unique nonzero shift_size
    has_sz0 <- any(df_n$sz_orig == 0, na.rm = TRUE)
    shift_levels_nonzero <- sort(unique(df_n$shift_size[df_n$sz_orig != 0]))
    panel_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                    paste0("shift=", shift_levels_nonzero))

    # Colors for psi (consistent across all panels in this n & reg)
    psi_all <- sort(unique(df_n$psi))
    colors <- hcl.colors(length(psi_all), palette = "Zissou 1")
    names(colors) <- as.character(psi_all)

    # Compute common ylim across panels (max_y)
    max_y <- 0
    for (pk in panel_keys) {
      df_panel <- if (pk == "pooled_sz0") {
        subset(df_n, sz_orig == 0)
      } else {
        sh <- as.numeric(sub("shift=", "", pk))
        subset(df_n, sz_orig != 0 & shift_size == sh)
      }
      for (psi_val in psi_all) {
        avg_counts <- avg_curve_for_subset(subset(df_panel, psi == psi_val))
        max_y <- max(max_y, avg_counts, na.rm = TRUE)
      }
    }
    if (!is.finite(max_y) || max_y <= 0) max_y <- 1

    # Layout
    K <- length(panel_keys)
    ncol <- ceiling(sqrt(K))
    nrow <- ceiling(K / ncol)
    png_w <- 900 * ncol
    png_h <- 700 * nrow
    out_path <- file.path(
      base_plot_folder,
      paste0("genotype_hist_avg_reg_", reg_lab, "_n_", n_val, "_gen_", target_gen, ".png")
    )
    png(out_path, width = png_w, height = png_h)
    op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3.5, 1) + 0.1)

    for (pk in panel_keys) {
      df_panel <- if (pk == "pooled_sz0") {
        subset(df_n, sz_orig == 0)
      } else {
        sh <- as.numeric(sub("shift=", "", pk))
        subset(df_n, sz_orig != 0 & shift_size == sh)
      }

      # Title text
      panel_title <- if (pk == "pooled_sz0") {
        paste0("reg = ", reg_lab, " | n = ", n_val, " | sz=0 pooled (data-driven)")
      } else {
        paste0("reg = ", reg_lab, " | n = ", n_val, " | ", pk)
      }

      plot(NULL,
           xlim = range(bin_breaks), ylim = c(0, max_y),
           xlab = "Genotype Value", ylab = "Proportion",
           main = panel_title)

      # Draw average curves per psi
      present_psis <- intersect(psi_all, sort(unique(df_panel$psi)))
      for (psi_val in present_psis) {
        avg_counts <- avg_curve_for_subset(subset(df_panel, psi == psi_val))
        if (all(is.na(avg_counts))) next
        lines(bin_mids, avg_counts, col = colors[as.character(psi_val)], lwd = 2)
      }

      # Gaussian overlay (scaled)
      Vs <- sqrt(2000)
      z_vals <- seq(min(bin_breaks), max(bin_breaks), length.out = 500)
      W_z <- exp(- (z_vals^2) / (2 * Vs))
      W_z_scaled <- W_z / max(W_z) * max_y
      lines(z_vals, W_z_scaled, col = "black", lwd = 2, lty = 2)

      legend("topright",
             legend = paste0("psi=", present_psis),
             col = colors[as.character(present_psis)],
             lty = 1, lwd = 2, cex = 0.8, bg = "white")
    }

    par(op)
    dev.off()
  }
}


```


# PLOT GENOTYPE OVER TIME
```{r}
# ---------------------------------------------
# Multi-panel by psi (GENOTYPE) with gen-specific bins
# Split by `reg` (values in `regimes`)
# ---------------------------------------------

# Derive a constant bin width from your reference breaks
bin_width <- (max(bin_breaks) - min(bin_breaks)) / num_bins
half_span <- (num_bins * bin_width) / 2

.compute_center <- function(df_block) {
  xcols <- grep("^X", names(df_block))
  if (!length(xcols) || nrow(df_block) == 0) return(NA_real_)
  vals <- suppressWarnings(as.numeric(unlist(df_block[, xcols])))
  vals <- vals[is.finite(vals)]
  if (!length(vals)) return(NA_real_)
  mean(vals)
}

.avg_hist_curve_with_breaks <- function(df_block, breaks, num_bins) {
  if (nrow(df_block) == 0) return(rep(NA_real_, num_bins))
  xcols <- grep("^X", names(df_block))
  reps  <- sort(unique(df_block$rep))
  if (!length(xcols) || !length(reps)) return(rep(NA_real_, num_bins))
  counts_mat <- matrix(NA_real_, nrow = length(reps), ncol = num_bins)
  for (i in seq_along(reps)) {
    row_i <- subset(df_block, rep == reps[i])[1, ]
    vals <- suppressWarnings(as.numeric(as.vector(unlist(row_i[, xcols]))))
    vals <- vals[is.finite(vals)]
    if (!length(vals)) next
    h <- hist(vals, breaks = breaks, plot = FALSE)
    if (sum(h$counts) == 0) next
    counts_mat[i, ] <- h$counts / sum(h$counts)
  }
  colMeans(counts_mat, na.rm = TRUE)
}

plot_genotype_multi_by_psi <- function(genotype_df,
                                       out_subfolder = "genotype_multi_by_psi_genbins",
                                       gen_seq = 10000:10005,
                                       num_bins = num_bins,
                                       bin_width = bin_width,
                                       half_span = half_span,
                                       regimes = regimes) {   # <--- added
  gen_cols <- hcl.colors(length(gen_seq), palette = "Berlin")
  names(gen_cols) <- as.character(gen_seq)

  if (!("sz_orig" %in% names(genotype_df))) {
    stop("genotype_df must include sz_orig (original sz) to pool data-driven shift runs.")
  }

  # Loop over regimes
  for (reg_lab in regimes) {
    df_reg <- subset(genotype_df, reg == reg_lab)
    if (nrow(df_reg) == 0) next

    # Separate subfolder per regime
    out_dir <- file.path(base_plot_folder, out_subfolder, paste0("reg_", reg_lab))
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

    for (n_val in sort(unique(df_reg$n))) {
      df_n <- subset(df_reg, n == n_val)
      if (nrow(df_n) == 0) next

      has_sz0 <- any(df_n$sz_orig == 0, na.rm = TRUE)
      nonzero_shifts <- sort(unique(df_n$shift_size[df_n$sz_orig != 0]))
      shift_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                      if (length(nonzero_shifts)) paste0("shift=", nonzero_shifts) else character(0))

      all_psis <- sort(unique(df_n$psi))

      for (sk in shift_keys) {
        df_group <- if (sk == "pooled_sz0") {
          subset(df_n, sz_orig == 0)
        } else {
          sh <- as.numeric(sub("shift=", "", sk))
          subset(df_n, sz_orig != 0 & shift_size == sh)
        }
        if (nrow(df_group) == 0) next

        # Common y-limit and x-range across all psi & gens
        max_y <- 0
        x_min <- Inf; x_max <- -Inf
        for (psi_val in all_psis) {
          df_p <- subset(df_group, psi == psi_val)
          if (nrow(df_p) == 0) next
          present_gens <- intersect(gen_seq, sort(unique(df_p$generation)))
          for (g in present_gens) {
            df_pg <- subset(df_p, generation == g)
            center <- .compute_center(df_pg)
            if (!is.finite(center)) next
            breaks_g <- seq(center - half_span, center + half_span, by = bin_width)
            avg_counts <- .avg_hist_curve_with_breaks(df_pg, breaks_g, num_bins)
            max_y <- max(max_y, avg_counts, na.rm = TRUE)
            x_min <- min(x_min, min(breaks_g), na.rm = TRUE)
            x_max <- max(x_max, max(breaks_g), na.rm = TRUE)
          }
        }
        if (!is.finite(max_y) || max_y <= 0) max_y <- 1
        if (!is.finite(x_min) || !is.finite(x_max) || x_min >= x_max) {
          x_min <- -half_span; x_max <- half_span
        }

        # Layout
        K <- length(all_psis)
        ncol <- ceiling(sqrt(K))
        nrow <- ceiling(K / ncol)
        png_w <- 900 * ncol
        png_h <- 700 * nrow

        # Overall title: do NOT average across psi
        title_suffix <- if (sk == "pooled_sz0") {
          "pooled sz=0 (per-psi avg shift shown in panels)"
        } else {
          sh <- as.numeric(sub("shift=", "", sk))
          paste0("shift_size = ", sh)
        }

        out_name <- paste0(
          "genotype_multi_by_psi_reg_", reg_lab, "_n_", n_val, "_",
          gsub("[ =]", "", title_suffix), ".png"
        )
        png(file.path(out_dir, out_name), width = png_w, height = png_h)
        op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3, 1) + 0.1, oma = c(0, 0, 3.5, 0))

        for (psi_val in all_psis) {
          df_p <- subset(df_group, psi == psi_val)

          # Panel title: if pooled sz=0, show per-psi mean shift_size
          panel_title <- if (sk == "pooled_sz0") {
            mpsi <- mean(df_p$shift_size, na.rm = TRUE)
            if (!is.finite(mpsi)) mpsi <- NA_real_
            paste0("psi = ", psi_val, " | avg shift_size = ",
                   if (is.na(mpsi)) "NA" else formatC(mpsi, format = "f", digits = 3))
          } else {
            paste0("psi = ", psi_val)
          }

          plot(NULL,
               xlim = c(x_min, x_max), ylim = c(0, max_y),
               xlab = "Genotype Value", ylab = "Proportion",
               main = panel_title)

          present_gens <- intersect(gen_seq, sort(unique(df_p$generation)))
          for (g in present_gens) {
            df_pg <- subset(df_p, generation == g)
            center <- .compute_center(df_pg)
            if (!is.finite(center)) next
            breaks_g <- seq(center - half_span, center + half_span, by = bin_width)
            mids_g   <- 0.5 * (breaks_g[-length(breaks_g)] + breaks_g[-1])
            avg_counts <- .avg_hist_curve_with_breaks(df_pg, breaks_g, num_bins)
            if (all(is.na(avg_counts))) next
            lines(mids_g, avg_counts, col = gen_cols[as.character(g)], lwd = 2)
          }

          legend("topright",
                 legend = paste0("gen ", present_gens),
                 col = gen_cols[as.character(present_gens)],
                 lty = 1, lwd = 2, cex = 0.8, bg = "white")
        }

        mtext(paste0("Genotype distributions (gen-specific bins) | reg = ", reg_lab,
                     " | n = ", n_val, " | ", title_suffix),
              outer = TRUE, cex = 1.1, line = 1.2)

        par(op)
        dev.off()
      }
    }
  }
}

# ---- Run ----
plot_genotype_multi_by_psi(
  genotype_df,
  out_subfolder = "genotype_multi_by_psi_genbins",
  gen_seq = 10000:10005,
  num_bins = num_bins,
  bin_width = bin_width,
  half_span = half_span,
  regimes = regimes   # e.g., c("ave","fit")
)

```


# load PHENOTYPE data
```{r}
pheno_folder <- file.path(base_data_folder, "phenotype")

# List all phenotype files
pheno_files <- list.files(
  path = pheno_folder,
  pattern = "\\.tsv$",
  full.names = TRUE,
  recursive = TRUE
)

# Parse metadata from phenotype file path
# Matches: phenotype/n_<n>_psi_<psi>_sz_<sz>_reg_<ave|fit>_<rep>.tsv
extract_pheno_metadata <- function(path) {
  m <- regexec("phenotype/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)\\.tsv", path)
  parts <- regmatches(path, m)[[1]]
  if (length(parts) != 6) {
    warning("Could not extract metadata from path: ", path, " parts: ", paste(parts, collapse=", "))
    return(data.frame())
  }
  data.frame(
    n   = as.integer(parts[2]),
    psi = as.numeric(parts[3]),
    sz  = as.integer(parts[4]),
    reg = parts[5],                 # "ave" or "fit"
    rep = as.integer(parts[6]),
    stringsAsFactors = FALSE
  )
}

# Get shift_size from companion "data" file (optimum at generation == 10001)
pheno_get_shift_from_data <- function(pheno_file, base_data_folder, opt_gen = 10001L) {
  # Mirror .../phenotype/... -> .../data/...
  data_path <- sub("/phenotype/", "/data/", pheno_file, fixed = TRUE)
  if (!file.exists(data_path)) {
    data_path <- file.path(base_data_folder, "data", basename(pheno_file))
  }
  if (!file.exists(data_path)) {
    warning("Data file not found for ", basename(pheno_file), " (looked for: ", data_path, ")")
    return(NA_real_)
  }

  d <- tryCatch(read.table(data_path, header = TRUE, sep = "\t"), error = function(e) NULL)
  if (is.null(d)) {
    warning("Failed to read data file: ", data_path)
    return(NA_real_)
  }
  if (!all(c("tick", "optimum") %in% names(d))) {
    warning("Missing 'tick' or 'optimum' in data file: ", data_path)
    return(NA_real_)
  }

  row10001 <- d[d$tick == opt_gen, , drop = FALSE]
  if (nrow(row10001) < 1) {
    warning("No row with tick == ", opt_gen, " in ", data_path)
    return(NA_real_)
  }

  val <- suppressWarnings(as.numeric(row10001$optimum[1]))
  if (!is.finite(val)) {
    warning("Non-numeric 'optimum' at generation ", opt_gen, " in ", data_path)
    return(NA_real_)
  }
  val
}

# -------- Load phenotype_df (no name collisions with genotype) --------
opt_gen <- 10001L
phenotype_df <- do.call(rbind, lapply(pheno_files, function(file) {
  df <- tryCatch(read.table(file, header = TRUE, sep = "\t"), error = function(e) NULL)
  if (is.null(df)) {
    warning("Failed to read phenotype file: ", file)
    return(NULL)
  }

  meta <- extract_pheno_metadata(file)
  if (nrow(meta) == 0) return(NULL)

  # Attach metadata (distinct columns; keep original sz for pooling)
  df$rep        <- meta$rep
  df$psi        <- meta$psi
  df$n          <- meta$n
  df$reg        <- meta$reg          # <- new column ("ave" or "fit")
  df$sz_orig    <- meta$sz
  df$shift_size <- if (meta$sz == 0L) pheno_get_shift_from_data(file, base_data_folder, opt_gen) else meta$sz
  df$file_id    <- basename(file)
  df
}))

```

# plot PHENOTYPE histogram 
```{r}
# Assuming you want to focus on a specific generation (e.g., 10100)
df <- subset(phenotype_df, generation == target_gen)

# Define histogram bins
num_bins <- 120
bin_breaks <- seq(-60, 60, length.out = num_bins + 1)
bin_mids <- 0.5 * (head(bin_breaks, -1) + tail(bin_breaks, -1))

# Color setup
psi_vals <- target_psi
colors <- rainbow(length(psi_vals))
names(colors) <- as.character(psi_vals)
```

## plot average PHENOTYPE histogram for each psi
```{r}
avg_curve_for_subset_pheno <- function(df_block, bin_breaks, bin_mids) {
  if (nrow(df_block) == 0) return(rep(NA_real_, length(bin_mids)))
  counts_mat <- matrix(NA_real_, nrow = nrow(df_block), ncol = length(bin_mids))
  x_cols <- grep("^X", names(df_block))
  for (i in seq_len(nrow(df_block))) {
    vals <- suppressWarnings(as.numeric(as.vector(unlist(df_block[i, x_cols]))))
    vals <- vals[is.finite(vals)]
    if (length(vals) == 0) next
    h <- hist(vals, breaks = bin_breaks, plot = FALSE)
    if (sum(h$counts) == 0) next
    counts_mat[i, ] <- h$counts / sum(h$counts)
  }
  colMeans(counts_mat, na.rm = TRUE)
}

# ---- Split by regime ----
for (reg_lab in regimes) {
  df_reg <- subset(phenotype_df, generation == target_gen & reg == reg_lab)
  if (nrow(df_reg) == 0) next

  # make a subfolder per regime
  out_dir <- file.path(base_plot_folder, paste0("reg_", reg_lab))
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  # Pool all sz_orig == 0 together; other panels by shift_size (nonzero)
  for (n_val in sort(unique(df_reg$n))) {
    df_n <- subset(df_reg, n == n_val)
    if (nrow(df_n) == 0) next

    has_sz0 <- any(df_n$sz_orig == 0, na.rm = TRUE)
    shift_levels_nonzero <- sort(unique(df_n$shift_size[df_n$sz_orig != 0]))
    panel_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                    paste0("shift=", shift_levels_nonzero))

    # Consistent psi colors across this n (within this reg)
    psi_all <- sort(unique(df_n$psi))
    psi_cols <- hcl.colors(length(psi_all), palette = "Zissou 1")
    names(psi_cols) <- as.character(psi_all)

    # Common ylim across panels
    max_y <- 0
    for (pk in panel_keys) {
      df_panel <- if (pk == "pooled_sz0") {
        subset(df_n, sz_orig == 0)
      } else {
        sh <- as.numeric(sub("shift=", "", pk))
        subset(df_n, sz_orig != 0 & shift_size == sh)
      }
      for (psi_val in psi_all) {
        avg_counts <- avg_curve_for_subset_pheno(subset(df_panel, psi == psi_val), bin_breaks, bin_mids)
        max_y <- max(max_y, avg_counts, na.rm = TRUE)
      }
    }
    if (!is.finite(max_y) || max_y <= 0) max_y <- 1

    # Layout
    K <- length(panel_keys)
    ncol <- ceiling(sqrt(K))
    nrow <- ceiling(K / ncol)
    png_w <- 900 * ncol
    png_h <- 700 * nrow

    out_path <- file.path(out_dir, paste0("phenotype_hist_avg_reg_", reg_lab,
                                          "_n_", n_val, "_gen_", target_gen, ".png"))
    png(out_path, width = png_w, height = png_h)
    op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3.5, 1) + 0.1)

    for (pk in panel_keys) {
      df_panel <- if (pk == "pooled_sz0") {
        subset(df_n, sz_orig == 0)
      } else {
        sh <- as.numeric(sub("shift=", "", pk))
        subset(df_n, sz_orig != 0 & shift_size == sh)
      }

      panel_title <- if (pk == "pooled_sz0") {
        paste0("reg = ", reg_lab, " | n = ", n_val, " | sz=0 pooled (data-driven)")
      } else {
        paste0("reg = ", reg_lab, " | n = ", n_val, " | ", pk)
      }

      plot(NULL,
           xlim = range(bin_breaks), ylim = c(0, max_y),
           xlab = "Phenotype Value", ylab = "Proportion",
           main = panel_title)

      present_psis <- intersect(psi_all, sort(unique(df_panel$psi)))
      for (psi_val in present_psis) {
        avg_counts <- avg_curve_for_subset_pheno(subset(df_panel, psi == psi_val), bin_breaks, bin_mids)
        if (all(is.na(avg_counts))) next
        lines(bin_mids, avg_counts, col = psi_cols[as.character(psi_val)], lwd = 2)
      }

      # Gaussian overlay
      Vs <- sqrt(2000)
      z_vals <- seq(min(bin_breaks), max(bin_breaks), length.out = 500)
      W_z <- exp(-(z_vals^2) / (2 * Vs))
      W_z_scaled <- W_z / max(W_z) * max_y
      lines(z_vals, W_z_scaled, lwd = 2, lty = 2)

      legend("topright",
             legend = paste0("psi=", present_psis),
             col = psi_cols[as.character(present_psis)],
             lty = 1, lwd = 2, cex = 0.8, bg = "white")
    }

    par(op)
    dev.off()
  }
}

```

# PLOT PHENOTYPE OVER TIME
```{r}
# ---- unchanged helpers above this line ----
# (bin_width, half_span, .ph_compute_center, .ph_avg_hist_with_breaks)

plot_phenotype_multi_by_psi <- function(phenotype_df,
                                        out_subfolder = "phenotype_multi_by_psi_genbins",
                                        gen_seq = 10000:10005,
                                        num_bins = num_bins,
                                        bin_width = bin_width,
                                        half_span = half_span,
                                        regimes = regimes) {

  gen_cols <- hcl.colors(length(gen_seq), palette = "Berlin")
  names(gen_cols) <- as.character(gen_seq)

  if (!("sz_orig" %in% names(phenotype_df))) {
    stop("phenotype_df must include sz_orig (original sz) to pool data-driven shift runs.")
  }
  if (!("reg" %in% names(phenotype_df))) {
    stop("phenotype_df must include 'reg' (e.g., 'ave' or 'fit').")
  }

  # Fixed Gaussian sd (requested)
  gauss_sd <- sqrt(2000)

  # Loop over regimes
  for (reg_lab in regimes) {
    df_reg <- subset(phenotype_df, reg == reg_lab)
    if (nrow(df_reg) == 0) next

    # Per-regime output dir
    out_dir <- file.path(base_plot_folder, out_subfolder, paste0("reg_", reg_lab))
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

    for (n_val in sort(unique(df_reg$n))) {
      df_n <- subset(df_reg, n == n_val)
      if (nrow(df_n) == 0) next

      has_sz0 <- any(df_n$sz_orig == 0, na.rm = TRUE)
      nonzero_shifts <- sort(unique(df_n$shift_size[df_n$sz_orig != 0]))
      shift_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                      if (length(nonzero_shifts)) paste0("shift=", nonzero_shifts) else character(0))

      all_psis <- sort(unique(df_n$psi))

      for (sk in shift_keys) {
        df_group <- if (sk == "pooled_sz0") {
          subset(df_n, sz_orig == 0)
        } else {
          sh <- as.numeric(sub("shift=", "", sk))
          subset(df_n, sz_orig != 0 & shift_size == sh)
        }
        if (nrow(df_group) == 0) next

        # Pass 1: common y-limit and x-range across all psi & gens
        max_y <- 0
        x_min <- Inf; x_max <- -Inf
        for (psi_val in all_psis) {
          df_p <- subset(df_group, psi == psi_val)
          if (nrow(df_p) == 0) next
          present_gens <- intersect(gen_seq, sort(unique(df_p$generation)))
          for (g in present_gens) {
            df_pg <- subset(df_p, generation == g)
            center <- .ph_compute_center(df_pg)
            if (!is.finite(center)) next
            breaks_g <- seq(center - half_span, center + half_span, by = bin_width)
            avg_counts <- .ph_avg_hist_with_breaks(df_pg, breaks_g, num_bins)
            max_y <- max(max_y, avg_counts, na.rm = TRUE)
            x_min <- min(x_min, min(breaks_g), na.rm = TRUE)
            x_max <- max(x_max, max(breaks_g), na.rm = TRUE)
          }
        }
        if (!is.finite(max_y) || max_y <= 0) max_y <- 1
        if (!is.finite(x_min) || !is.finite(x_max) || x_min >= x_max) {
          x_min <- -half_span; x_max <- half_span
        }

        # Layout
        K <- length(all_psis)
        ncol <- ceiling(sqrt(K))
        nrow <- ceiling(K / ncol)
        png_w <- 900 * ncol
        png_h <- 700 * nrow

        title_suffix <- if (sk == "pooled_sz0") {
          "pooled sz=0 (per-psi avg shift shown in panels)"
        } else {
          sh <- as.numeric(sub("shift=", "", sk))
          paste0("shift_size = ", sh)
        }

        out_name <- paste0("phenotype_multi_by_psi_reg_", reg_lab,
                           "_n_", n_val, "_", gsub("[ =]", "", title_suffix), ".png")
        png(file.path(out_dir, out_name), width = png_w, height = png_h)
        op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3, 1) + 0.1, oma = c(0, 0, 3.5, 0))

        for (psi_val in all_psis) {
          df_p <- subset(df_group, psi == psi_val)

          # Panel title: if pooled sz=0, show per-psi mean shift_size
          panel_title <- if (sk == "pooled_sz0") {
            mpsi <- mean(df_p$shift_size, na.rm = TRUE)
            if (!is.finite(mpsi)) mpsi <- NA_real_
            paste0("psi = ", psi_val, " | avg shift_size = ",
                   if (is.na(mpsi)) "NA" else formatC(mpsi, format = "f", digits = 3))
          } else {
            paste0("psi = ", psi_val)
          }

          plot(NULL,
               xlim = c(x_min, x_max), ylim = c(0, max_y),
               xlab = "Phenotype Value", ylab = "Proportion",
               main = panel_title)

          present_gens <- intersect(gen_seq, sort(unique(df_p$generation)))
          for (g in present_gens) {
            df_pg <- subset(df_p, generation == g)
            center <- .ph_compute_center(df_pg)
            if (!is.finite(center)) next
            breaks_g <- seq(center - half_span, center + half_span, by = bin_width)
            mids_g   <- 0.5 * (breaks_g[-length(breaks_g)] + breaks_g[-1])
            avg_counts <- .ph_avg_hist_with_breaks(df_pg, breaks_g, num_bins)
            if (all(is.na(avg_counts))) next
            lines(mids_g, avg_counts, col = gen_cols[as.character(g)], lwd = 2)
          }

          # ---- NEW: overlay Gaussian curve on same y-scale (probability per bin) ----
          # Choose optimum (mu):
          # - fixed-shift panels: the shift value in the key (equals sz_orig when sz_orig != 0)
          # - pooled sz=0 panels: the per-psi mean shift_size (data-driven optimum)
          mu <- if (sk == "pooled_sz0") {
            mean(df_p$shift_size, na.rm = TRUE)
          } else {
            as.numeric(sub("shift=", "", sk))
          }
          if (!is.finite(mu)) mu <- 0  # fallback

          # Draw a smooth Gaussian line across the panel, scaled by bin_width
          xgrid <- seq(x_min, x_max, length.out = 1000)
          ygrid <- dnorm(xgrid, mean = mu, sd = gauss_sd) * bin_width
          lines(xgrid, ygrid, lty = 2, lwd = 2)  # default color so it stands out against gen lines

          # Legend (include Gaussian entry)
          leg_labels <- c(paste0("gen ", present_gens), paste0("Gaussian (μ=", formatC(mu, digits = 3, format = "f"),
                                                               ", σ=√2000)"))
          leg_cols   <- c(gen_cols[as.character(present_gens)], par("fg"))
          leg_lty    <- c(rep(1, length(present_gens)), 2)
          legend("topright", legend = leg_labels, col = leg_cols,
                 lty = leg_lty, lwd = 2, cex = 0.8, bg = "white")
        }

        mtext(paste0("Phenotype distributions (gen-specific bins) | reg = ", reg_lab,
                     " | n = ", n_val, " | ", title_suffix),
              outer = TRUE, cex = 1.1, line = 1.2)

        par(op)
        dev.off()
      }
    }
  }
}

# ---- Run ----
plot_phenotype_multi_by_psi(
  phenotype_df,
  out_subfolder = "phenotype_multi_by_psi_genbins",
  gen_seq = 10000:10005,
  num_bins = num_bins,
  bin_width = bin_width,
  half_span = half_span,
  regimes = regimes   # e.g., c("ave","fit")
)
```

# load SUMMARY STATS data
```{r}
# Top-level folder
folder <- file.path(base_data_folder, "data")

# List all .tsv files recursively
file_list <- list.files(
  path = folder,
  pattern = "\\.tsv$",
  full.names = TRUE,
  recursive = TRUE
)

# Extract metadata from filename:
#   data/n_<n>/psi_<psi>/sz_<shift_size>/_reg_<ave|fit>_<rep>.tsv
extract_sum_metadata <- function(path) {
  rx <- "data/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)\\.tsv"
  m  <- regexec(rx, path)
  p  <- regmatches(path, m)[[1]]

  if (length(p) != 6) {
    warning("Could not extract metadata from path: ", path)
    return(data.frame())
  }

  data.frame(
    n          = as.integer(p[2]),
    psi        = as.numeric(p[3]),
    shift_size = as.integer(p[4]),
    reg        = p[5],                 # "ave" or "fit"
    rep        = as.integer(p[6]),
    file_id    = basename(path),
    stringsAsFactors = FALSE
  )
}

# Read and combine all files
summary_stat_df <- do.call(rbind, lapply(file_list, function(file) {
  df <- tryCatch(read.table(file, header = TRUE, sep = "\t", check.names = FALSE),
                 error = function(e) NULL)
  if (is.null(df)) {
    warning("Failed to read: ", file)
    return(NULL)
  }

  meta <- extract_sum_metadata(file)
  if (nrow(meta) == 0) return(NULL)

  # Attach metadata columns
  df$n          <- meta$n
  df$psi        <- meta$psi
  df$shift_size <- meta$shift_size
  df$reg        <- meta$reg
  df$rep        <- meta$rep
  df$file_id    <- meta$file_id

  df
}))

```

## plot SUMMARY STATS (z_sd over n)
```{r}
# ------------------------------------------------------
# Summary stats: z_sd vs n, panels = psi, per shift group
# Base R only
# ------------------------------------------------------

# Tick window to average z_sd per (n, psi, rep)
tick_seq <- 10000:10005

rep_col  <- adjustcolor("gray30", alpha.f = 0.35)  # replicate points
mean_col <- "black"                                # across-rep mean

# Per-psi pooled shift label (prefer 'optimum' at tick==10001 if present)
.per_psi_avg_shift <- function(df_panel) {
  if ("optimum" %in% names(df_panel)) {
    v <- df_panel$optimum[df_panel$tick == 10001]
    v <- v[is.finite(v)]
    if (length(v)) return(mean(v))
  }
  v <- df_panel$shift_size
  v <- v[is.finite(v)]
  if (!length(v)) return(NA_real_)
  mean(v)
}

# For a given (shift group × psi), compute:
# - x_n: sorted n values present
# - rep_values: list, one numeric vector per n (replicate-avg z_sd across tick_seq)
# - mean_values: across-rep means at each n
.compute_zsd_vs_n <- function(df_panel, tick_seq) {
  needed <- c("z_sd","n","psi","rep","tick")
  if (!all(needed %in% names(df_panel))) {
    return(list(x_n = numeric(0), rep_values = list(), mean_values = numeric(0)))
  }
  d <- df_panel
  d$n    <- suppressWarnings(as.integer(d$n))
  d$rep  <- suppressWarnings(as.integer(d$rep))
  d$tick <- suppressWarnings(as.integer(d$tick))
  d$z_sd <- suppressWarnings(as.numeric(d$z_sd))
  d <- d[is.finite(d$z_sd) & is.finite(d$n) & is.finite(d$rep) & d$tick %in% tick_seq, , drop = FALSE]
  if (!nrow(d)) return(list(x_n = numeric(0), rep_values = list(), mean_values = numeric(0)))

  # Average z_sd across ticks per (n, rep)
  agg <- aggregate(z_sd ~ n + rep, data = d, FUN = mean)

  x_n <- sort(unique(agg$n))
  rep_values <- lapply(x_n, function(nv) {
    vals <- agg$z_sd[agg$n == nv]
    vals[is.finite(vals)]
  })

  mean_values <- vapply(rep_values, function(v) {
    if (!length(v)) return(NA_real_)
    mean(v, na.rm = TRUE)
  }, numeric(1))

  list(x_n = x_n, rep_values = rep_values, mean_values = mean_values)
}

plot_zsd_vs_n_by_psi <- function(summary_stat_df,
                                 out_subfolder = "summary_zsd_vs_n_by_psi",
                                 tick_seq = tick_seq) {
  out_dir <- file.path(base_plot_folder, out_subfolder)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  has_sz_orig <- "sz_orig" %in% names(summary_stat_df)
  all_psis <- sort(unique(summary_stat_df$psi))

  # One PNG per shift group: pooled 0 and each non-zero
  for (shift_grouping in c("pooled", "fixed")) {

    # Build groups to render
    if (shift_grouping == "pooled") {
      df_pool <- if (has_sz_orig) subset(summary_stat_df, sz_orig == 0)
                 else subset(summary_stat_df, shift_size == 0)
      if (nrow(df_pool) == 0) next
      group_list <- list(list(key = "pooled_sz0", df = df_pool))
    } else {
      if (has_sz_orig) {
        fixed_vals <- sort(unique(summary_stat_df$shift_size[summary_stat_df$sz_orig != 0]))
        groups <- lapply(fixed_vals, function(sh) {
          list(key = paste0("shift=", sh),
               df = subset(summary_stat_df,
                           ifelse(is.na(sz_orig), FALSE, sz_orig != 0) & shift_size == sh))
        })
      } else {
        fixed_vals <- sort(unique(summary_stat_df$shift_size[summary_stat_df$shift_size != 0]))
        groups <- lapply(fixed_vals, function(sh) {
          list(key = paste0("shift=", sh),
               df = subset(summary_stat_df, shift_size == sh))
        })
      }
      group_list <- Filter(function(g) nrow(g$df) > 0, groups)
      if (!length(group_list)) next
    }

    # Render each group
    for (ginfo in group_list) {
      df_group <- ginfo$df
      skey     <- ginfo$key

      # y-range across all psi panels
      y_min <- Inf; y_max <- -Inf
      x_all <- integer(0)
      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val)
        s <- .compute_zsd_vs_n(df_p, tick_seq)
        if (length(s$mean_values)) {
          y_all <- c(unlist(s$rep_values, use.names = FALSE), s$mean_values)
          y_all <- y_all[is.finite(y_all)]
          if (length(y_all)) {
            y_min <- min(y_min, y_all)
            y_max <- max(y_max, y_all)
          }
        }
        x_all <- sort(unique(c(x_all, s$x_n)))
      }
      if (!is.finite(y_min) || !is.finite(y_max) || y_min == y_max) { y_min <- 0; y_max <- 1 }
      if (!length(x_all)) next

      # Layout
      K <- length(all_psis)
      ncol <- ceiling(sqrt(K))
      nrow <- ceiling(K / ncol)
      png_w <- 900 * ncol
      png_h <- 700 * nrow

      title_suffix <- if (skey == "pooled_sz0") {
        paste0("pooled sz=0 (per-psi avg shift shown in panels; ticks ",
               min(tick_seq), "–", max(tick_seq), ")")
      } else {
        sh <- as.numeric(sub("shift=", "", skey))
        paste0("shift_size = ", sh, " (ticks ", min(tick_seq), "–", max(tick_seq), ")")
      }

      out_name <- paste0("summary_zsd_vs_n_", gsub("[ =]", "", skey), ".png")
      png(file.path(out_dir, out_name), width = png_w, height = png_h)
      op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3, 1) + 0.1, oma = c(0, 0, 3.5, 0))

      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val)

        panel_title <- if (skey == "pooled_sz0") {
          mpsi <- .per_psi_avg_shift(df_p)
          paste0("psi = ", psi_val, " | avg shift_size = ",
                 if (is.finite(mpsi)) formatC(mpsi, format = "f", digits = 3) else "NA")
        } else {
          paste0("psi = ", psi_val)
        }

        s <- .compute_zsd_vs_n(df_p, tick_seq)
        x_n <- s$x_n

        plot(NA, xlim = range(x_all), ylim = c(y_min, y_max),
             xlab = "n", ylab = "z_sd (avg over ticks)", main = panel_title,
             xaxt = "n")
        axis(1, at = sort(unique(x_all)))

        # Replicate points per n (jitter to reduce overplot)
        if (length(s$rep_values)) {
          for (i in seq_along(x_n)) {
            nv <- x_n[i]
            yy <- s$rep_values[[i]]
            if (!length(yy)) next
            xx <- jitter(rep(nv, length(yy)), amount = 0.15)
            points(xx, yy, pch = 16, cex = 0.6, col = rep_col)
          }
        }

        # Across-rep mean across n
        if (length(s$mean_values) && length(x_n)) {
          ord <- order(x_n)
          lines(x_n[ord], s$mean_values[ord], col = mean_col, lwd = 2)
          points(x_n[ord], s$mean_values[ord], pch = 16, cex = 0.9, col = mean_col)
        }

        legend("topleft", inset = 0.02,
               legend = c("Replicate mean (over ticks)", "Across-rep mean"),
               col = c(rep_col, mean_col), lty = c(0,1), pch = c(16,16),
               pt.cex = c(0.6,0.9), lwd = c(NA,2), bty = "n", cex = 0.85)
      }

      mtext(paste0("z_sd vs n (avg over ticks ", min(tick_seq), "–", max(tick_seq), ") | ", title_suffix),
            outer = TRUE, cex = 1.1, line = 1.2)

      par(op)
      dev.off()
    }
  }
}

# ---- Run ----
plot_zsd_vs_n_by_psi(summary_stat_df, out_subfolder = "summary_zsd_vs_n_by_psi", tick_seq = 10000:10005)

```

### a_sd over n averaged
```{r}
# ------------------------------------------------------
# Summary stats: a_sd vs n, panels = psi, per shift group
# Base R only
# ------------------------------------------------------

# Tick window used to average a_sd per (n, psi, rep)
tick_seq <- 10000:10005

rep_col  <- adjustcolor("gray30", alpha.f = 0.35)  # replicate points
mean_col <- "black"                                # across-rep mean

# Per-psi pooled shift label (prefer 'optimum' at tick==10001 if present)
.per_psi_avg_shift <- function(df_panel) {
  if ("optimum" %in% names(df_panel)) {
    v <- df_panel$optimum[df_panel$tick == 10001]
    v <- v[is.finite(v)]
    if (length(v)) return(mean(v))
  }
  v <- df_panel$shift_size
  v <- v[is.finite(v)]
  if (!length(v)) return(NA_real_)
  mean(v)
}

# For a given (shift group × psi), compute:
# - x_n: sorted n values present
# - rep_values: list, one numeric vector per n (replicate-averaged a_sd across tick_seq)
# - mean_values: numeric vector of across-rep means at each n
.compute_asd_vs_n <- function(df_panel, tick_seq) {
  needed <- c("a_sd","n","psi","rep","tick")
  if (!all(needed %in% names(df_panel))) {
    return(list(x_n = numeric(0), rep_values = list(), mean_values = numeric(0)))
  }
  # Coerce types, filter tick window
  d <- df_panel
  d$n    <- suppressWarnings(as.integer(d$n))
  d$rep  <- suppressWarnings(as.integer(d$rep))
  d$tick <- suppressWarnings(as.integer(d$tick))
  d$a_sd <- suppressWarnings(as.numeric(d$a_sd))
  d <- d[is.finite(d$a_sd) & is.finite(d$n) & is.finite(d$rep) & d$tick %in% tick_seq, , drop = FALSE]
  if (!nrow(d)) return(list(x_n = numeric(0), rep_values = list(), mean_values = numeric(0)))

  # Average a_sd across tick window per (n, rep)
  agg <- aggregate(a_sd ~ n + rep, data = d, FUN = mean)

  x_n <- sort(unique(agg$n))
  rep_values <- lapply(x_n, function(nv) {
    vals <- agg$a_sd[agg$n == nv]
    vals[is.finite(vals)]
  })

  mean_values <- vapply(rep_values, function(v) {
    if (!length(v)) return(NA_real_)
    mean(v, na.rm = TRUE)
  }, numeric(1))

  list(x_n = x_n, rep_values = rep_values, mean_values = mean_values)
}

plot_asd_vs_n_by_psi <- function(summary_stat_df,
                                 out_subfolder = "summary_asd_vs_n_by_psi",
                                 tick_seq = tick_seq) {
  out_dir <- file.path(base_plot_folder, out_subfolder)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  has_sz_orig <- "sz_orig" %in% names(summary_stat_df)
  all_psis <- sort(unique(summary_stat_df$psi))

  # We produce one PNG per shift group (pooled 0 and each non-zero)
  # spanning ALL n values for that group
  # (so we do NOT split by n here; x-axis is n)
  for (shift_grouping in c("pooled", "fixed")) {

    # Build list of groups to render
    if (shift_grouping == "pooled") {
      if (has_sz_orig) {
        df_pool <- subset(summary_stat_df, sz_orig == 0)
      } else {
        df_pool <- subset(summary_stat_df, shift_size == 0)
      }
      if (nrow(df_pool) == 0) next
      group_list <- list(list(key = "pooled_sz0", df = df_pool))
    } else {
      if (has_sz_orig) {
        fixed_vals <- sort(unique(summary_stat_df$shift_size[summary_stat_df$sz_orig != 0]))
        groups <- lapply(fixed_vals, function(sh) {
          list(key = paste0("shift=", sh),
               df = subset(summary_stat_df,
                           ifelse(is.na(sz_orig), FALSE, sz_orig != 0) & shift_size == sh))
        })
      } else {
        fixed_vals <- sort(unique(summary_stat_df$shift_size[summary_stat_df$shift_size != 0]))
        groups <- lapply(fixed_vals, function(sh) {
          list(key = paste0("shift=", sh),
               df = subset(summary_stat_df, shift_size == sh))
        })
      }
      # drop empties
      group_list <- Filter(function(g) nrow(g$df) > 0, groups)
      if (!length(group_list)) next
    }

    # Render each group to its own PNG
    for (ginfo in group_list) {
      df_group <- ginfo$df
      skey     <- ginfo$key

      # y-range across all psi panels in this group
      y_min <- Inf; y_max <- -Inf
      x_all <- integer(0)
      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val)
        s <- .compute_asd_vs_n(df_p, tick_seq)
        if (length(s$mean_values)) {
          # Collect y-range from both replicate points and means
          y_all <- c(unlist(s$rep_values, use.names = FALSE), s$mean_values)
          y_all <- y_all[is.finite(y_all)]
          if (length(y_all)) {
            y_min <- min(y_min, y_all)
            y_max <- max(y_max, y_all)
          }
        }
        x_all <- sort(unique(c(x_all, s$x_n)))
      }
      if (!is.finite(y_min) || !is.finite(y_max) || y_min == y_max) { y_min <- 0; y_max <- 1 }
      if (!length(x_all)) next

      # layout
      K <- length(all_psis)
      ncol <- ceiling(sqrt(K))
      nrow <- ceiling(K / ncol)
      png_w <- 900 * ncol
      png_h <- 700 * nrow

      title_suffix <- if (skey == "pooled_sz0") {
        paste0("pooled sz=0 (per-psi avg shift shown in panels; ticks ",
               min(tick_seq), "–", max(tick_seq), ")")
      } else {
        sh <- as.numeric(sub("shift=", "", skey))
        paste0("shift_size = ", sh, " (ticks ", min(tick_seq), "–", max(tick_seq), ")")
      }

      out_name <- paste0("summary_asd_vs_n_", gsub("[ =]", "", skey), ".png")
      png(file.path(out_dir, out_name), width = png_w, height = png_h)
      op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3, 1) + 0.1, oma = c(0, 0, 3.5, 0))

      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val)

        # Per-panel title
        panel_title <- if (skey == "pooled_sz0") {
          mpsi <- .per_psi_avg_shift(df_p)
          paste0("psi = ", psi_val, " | avg shift_size = ",
                 if (is.finite(mpsi)) formatC(mpsi, format = "f", digits = 3) else "NA")
        } else {
          paste0("psi = ", psi_val)
        }

        # Compute series for this psi
        s <- .compute_asd_vs_n(df_p, tick_seq)
        x_n <- s$x_n
        plot(NA, xlim = range(x_all), ylim = c(y_min, y_max),
             xlab = "n", ylab = "a_sd (avg over ticks)", main = panel_title,
             xaxt = "n")
        axis(1, at = sort(unique(x_all)))

        # Replicate points at each n (slight jitter to reduce overplot)
        if (length(s$rep_values)) {
          for (i in seq_along(x_n)) {
            nv <- x_n[i]
            yy <- s$rep_values[[i]]
            if (!length(yy)) next
            xx <- jitter(rep(nv, length(yy)), amount = 0.15)
            points(xx, yy, pch = 16, cex = 0.6, col = rep_col)
          }
        }

        # Across-rep mean as a line across n
        if (length(s$mean_values) && length(x_n)) {
          ord <- order(x_n)
          lines(x_n[ord], s$mean_values[ord], col = mean_col, lwd = 2)
          points(x_n[ord], s$mean_values[ord], pch = 16, cex = 0.9, col = mean_col)
        }

        legend("topleft", inset = 0.02,
               legend = c("Replicate mean (over ticks)", "Across-rep mean"),
               col = c(rep_col, mean_col), lty = c(0,1), pch = c(16,16),
               pt.cex = c(0.6,0.9), lwd = c(NA,2), bty = "n", cex = 0.85)
      }

      mtext(paste0("a_sd vs n (avg over ticks ", min(tick_seq), "–", max(tick_seq), ") | ", title_suffix),
            outer = TRUE, cex = 1.1, line = 1.2)

      par(op)
      dev.off()
    }
  }
}

# ---- Run it ----
plot_asd_vs_n_by_psi(summary_stat_df, out_subfolder = "summary_asd_vs_n_by_psi", tick_seq = 10000:10005)

```

## plot dist_norm over TIME
```{r}
# ------------------------------------------------------
# Summary stats multipanel (x = tick, y = dist_norm * Va0_sqrt)
# Panels = psi; one PNG per (n, shift-group), base R only
# ------------------------------------------------------

tick_seq <- 10000:10100
rep_col  <- adjustcolor("gray30", alpha.f = 0.35)
mean_col <- "black"

# Per-psi pooled shift label: prefer 'optimum' at tick==10001 if present
.per_psi_avg_shift <- function(df_panel) {
  if ("optimum" %in% names(df_panel)) {
    v <- df_panel$optimum[df_panel$tick == 10001]
    v <- v[is.finite(v)]
    if (length(v)) return(mean(v))
  }
  v <- df_panel$shift_size
  v <- v[is.finite(v)]
  if (!length(v)) return(NA_real_)
  mean(v)
}

# Build per-rep series over tick_seq and across-rep mean
.compute_series <- function(df_panel, tick_seq) {
  needed <- c("dist_norm","Va0_sqrt","tick","rep")
  if (!all(needed %in% names(df_panel))) {
    return(list(reps = list(), mean = rep(NA_real_, length(tick_seq)), present_ticks = integer(0)))
  }
  df_sub <- df_panel
  df_sub$y <- suppressWarnings(as.numeric(df_sub$dist_norm) * as.numeric(df_sub$Va0_sqrt))
  df_sub$tick <- suppressWarnings(as.integer(df_sub$tick))
  df_sub$rep  <- suppressWarnings(as.integer(df_sub$rep))
  df_sub <- df_sub[is.finite(df_sub$y) & is.finite(df_sub$tick) & is.finite(df_sub$rep), , drop = FALSE]
  if (!nrow(df_sub)) return(list(reps = list(), mean = rep(NA_real_, length(tick_seq)), present_ticks = integer(0)))

  # If duplicates per (tick, rep), average them
  agg <- aggregate(y ~ tick + rep, data = df_sub, FUN = mean)

  reps <- sort(unique(agg$rep))
  rep_series <- lapply(reps, function(r) {
    yv <- rep(NA_real_, length(tick_seq))
    rows <- agg[agg$rep == r & agg$tick %in% tick_seq, , drop = FALSE]
    if (nrow(rows)) {
      idx <- match(rows$tick, tick_seq)
      yv[idx] <- rows$y
    }
    yv
  })
  names(rep_series) <- as.character(reps)

  if (length(rep_series)) {
    mat <- do.call(rbind, rep_series)  # reps x ticks
    mean_series  <- apply(mat, 2L, function(col) mean(col, na.rm = TRUE))
    present_ticks <- tick_seq[colSums(!is.na(mat)) > 0]
  } else {
    mean_series <- rep(NA_real_, length(tick_seq))
    present_ticks <- integer(0)
  }

  list(reps = rep_series, mean = mean_series, present_ticks = present_ticks)
}

plot_summary_multi_by_psi_ticks <- function(summary_stat_df,
                                            out_subfolder = "summary_multi_by_psi") {
  out_dir <- file.path(base_plot_folder, out_subfolder)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  has_sz_orig <- "sz_orig" %in% names(summary_stat_df)

  for (n_val in sort(unique(summary_stat_df$n))) {
    df_n <- subset(summary_stat_df, n == n_val)
    if (nrow(df_n) == 0) next

    # Shift groups: pooled sz=0 + each nonzero shift
    if (has_sz_orig) {
      has_sz0 <- any(df_n$sz_orig == 0, na.rm = TRUE)
      nonzero <- sort(unique(df_n$shift_size[df_n$sz_orig != 0]))
      shift_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                      if (length(nonzero)) paste0("shift=", nonzero) else character(0))
    } else {
      has_sz0 <- any(df_n$shift_size == 0, na.rm = TRUE)
      nonzero <- sort(unique(df_n$shift_size[df_n$shift_size != 0]))
      shift_keys <- c(if (has_sz0) "pooled_sz0" else character(0),
                      if (length(nonzero)) paste0("shift=", nonzero) else character(0))
    }

    all_psis <- sort(unique(df_n$psi))

    for (sk in shift_keys) {
      df_group <- if (sk == "pooled_sz0") {
        if (has_sz_orig) subset(df_n, sz_orig == 0) else subset(df_n, shift_size == 0)
      } else {
        sh <- as.numeric(sub("shift=", "", sk))
        if (has_sz_orig) subset(df_n, ifelse(is.na(sz_orig), FALSE, sz_orig != 0) & shift_size == sh)
        else subset(df_n, shift_size == sh)
      }
      if (nrow(df_group) == 0) next

      # Y-limits across all psi panels
      y_min <- Inf; y_max <- -Inf
      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val & tick %in% tick_seq)
        s <- .compute_series(df_p, tick_seq)
        y_all <- c(unlist(s$reps, use.names = FALSE), s$mean)
        y_all <- y_all[is.finite(y_all)]
        if (length(y_all)) {
          y_min <- min(y_min, y_all)
          y_max <- max(y_max, y_all)
        }
      }
      if (!is.finite(y_min) || !is.finite(y_max) || y_min == y_max) {
        y_min <- 0; y_max <- 1
      }

      # Layout
      K <- length(all_psis)
      ncol <- ceiling(sqrt(K))
      nrow <- ceiling(K / ncol)
      png_w <- 900 * ncol
      png_h <- 700 * nrow

      # Title
      title_suffix <- if (sk == "pooled_sz0") {
        "pooled sz=0 (per-psi avg shift shown in panels)"
      } else {
        sh <- as.numeric(sub("shift=", "", sk))
        paste0("shift_size = ", sh)
      }

      out_name <- paste0("summary_multi_by_psi_ticks_n_", n_val, "_", gsub("[ =]", "", title_suffix), ".png")
      png(file.path(out_dir, out_name), width = png_w, height = png_h)
      op <- par(mfrow = c(nrow, ncol), mar = c(4, 4, 3, 1) + 0.1, oma = c(0, 0, 3.5, 0))

      for (psi_val in all_psis) {
        df_p <- subset(df_group, psi == psi_val & tick %in% tick_seq)

        panel_title <- if (sk == "pooled_sz0") {
          mpsi <- .per_psi_avg_shift(df_p)
          paste0("psi = ", psi_val, " | avg shift_size = ",
                 if (is.finite(mpsi)) formatC(mpsi, format = "f", digits = 3) else "NA")
        } else {
          paste0("psi = ", psi_val)
        }

        plot(NA, xlim = range(tick_seq), ylim = c(y_min, y_max),
             xlab = "Tick", ylab = "dist_norm × Va0_sqrt", main = panel_title,
             xaxt = "n")
        axis(1, at = tick_seq)

        s <- .compute_series(df_p, tick_seq)

        # Rep series (gray)
        if (length(s$reps)) {
          for (r in seq_along(s$reps)) {
            lines(tick_seq, s$reps[[r]], col = rep_col, lwd = 1)
            points(tick_seq, s$reps[[r]], pch = 16, cex = 0.5, col = rep_col)
          }
        }
        # Across-rep mean (black)
        lines(tick_seq, s$mean, col = mean_col, lwd = 2)
        points(tick_seq, s$mean, pch = 16, cex = 0.9, col = mean_col)

        legend("topleft", inset = 0.02,
               legend = c("Replicate", "Across-rep mean"),
               col = c(rep_col, mean_col), lty = 1, lwd = c(1,2), pch = c(16,16),
               pt.cex = c(0.5,0.9), bty = "n", cex = 0.85)
      }

      mtext(paste0("Summary: dist_norm × Va0_sqrt over ticks | n = ", n_val, " | ", title_suffix),
            outer = TRUE, cex = 1.1, line = 1.2)

      par(op)
      dev.off()
    }
  }
}

# ---- Run ----
plot_summary_multi_by_psi_ticks(summary_stat_df, out_subfolder = "summary_multi_by_psi_ticks")
```

## process ind data
```{r}
# -------------------------
# BASE R IND-EFFECTS LOADER (fixed)
# -------------------------

# ---- CONFIG ----
folder      <- file.path(base_data_folder, "ind")
gen_filter  <- "10000"   # encoded in filenames to keep
target_strs <- sprintf("psi_%s", formatC(target_psi, format = "f", digits = 1))  # e.g., psi_-0.3

# ---- HELPERS ----

# Filename: ind/n_{n}_psi_{psi}_sz_{shift}_reg_{ave|fit}_{rep}_gen{gen_filter}.tsv
extract_effect_metadata <- function(path) {
  rx <- paste0("ind/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)_gen", gen_filter, "\\.tsv")
  m  <- regexec(rx, path)
  p  <- regmatches(path, m)[[1]]
  if (length(p) != 6) {
    warning("Could not extract metadata from path: ", path, " | parts: ", paste(p, collapse = ", "))
    return(NULL)
  }
  list(
    n          = as.integer(p[2]),
    psi        = as.numeric(p[3]),
    shift_size = as.integer(p[4]),
    reg        = p[5],               # "ave" or "fit"
    rep        = as.integer(p[6]),
    file_id    = basename(path)
  )
}

# Read variable-length TSV with base R; no headers; pad shorter rows
read_variable_tsv_base <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)
  if (!length(lines)) return(NULL)
  split_lines <- strsplit(lines, "\t", fixed = TRUE)
  max_len <- max(vapply(split_lines, length, integer(1)))
  padded <- lapply(split_lines, function(row) { length(row) <- max_len; row })
  df <- as.data.frame(do.call(rbind, padded), stringsAsFactors = FALSE)
  if (ncol(df) < 2) {
    warning("File has fewer than 2 columns: ", file_path)
    return(NULL)
  }
  names(df)[1:2] <- c("generation", "individual_id")
  suppressWarnings({
    df$generation    <- as.integer(df$generation)
    df$individual_id <- as.integer(df$individual_id)
  })
  df
}

# Parse tokens like "mutID:effect" or just "effect" -> numeric effect
tokens_to_effects <- function(x) {
  if (is.null(x) || !length(x)) return(numeric(0))
  x <- x[!is.na(x) & nzchar(x)]
  if (!length(x)) return(numeric(0))
  fx <- sub("^.*?:", "", x, perl = TRUE)   # keep substring after last ":" if present
  suppressWarnings(as.numeric(fx))
}

# ---- LIST FILES ----
file_list <- list.files(
  path = folder,
  pattern = paste0("_gen", gen_filter, "\\.tsv$"),
  full.names = TRUE,
  recursive = TRUE
)

# filter by psi string (1 decimal in filenames)
file_list <- Filter(function(f) any(sapply(target_strs, function(p) grepl(p, f, fixed = TRUE))), file_list)

# ---- READ ALL FILES ----
df_list <- list()
for (file in file_list) {
  raw_df <- read_variable_tsv_base(file)
  if (is.null(raw_df) || !nrow(raw_df)) next

  meta <- extract_effect_metadata(file)
  if (is.null(meta)) next

  # attach metadata (vectorized per row)
  raw_df$rep        <- meta$rep
  raw_df$psi        <- meta$psi
  raw_df$n          <- meta$n
  raw_df$shift_size <- meta$shift_size
  raw_df$reg        <- meta$reg
  raw_df$file_id    <- meta$file_id

  df_list[[length(df_list) + 1]] <- raw_df
}
if (!length(df_list)) stop("No files loaded. Check folder/filters/regex.")

# ---- PROCESS: list-column 'effects' (robust against varying column counts) ----
effects_rows <- lapply(df_list, function(df) {
  # token columns = everything except the known core columns
  core <- c("generation","individual_id","rep","psi","n","shift_size","reg","file_id")
  tok_idx <- which(!(names(df) %in% core))

  eff_list <- if (!length(tok_idx)) {
    replicate(nrow(df), numeric(0), simplify = FALSE)
  } else {
    tmp <- apply(df[, tok_idx, drop = FALSE], 1L, tokens_to_effects)
    if (!is.list(tmp)) as.list(tmp) else tmp
  }

  out <- data.frame(
    generation    = df$generation,
    individual_id = df$individual_id,
    rep           = df$rep,
    psi           = as.numeric(df$psi),
    n             = as.integer(df$n),
    shift_size    = as.integer(df$shift_size),
    reg           = df$reg,
    file_id       = df$file_id,
    stringsAsFactors = FALSE
  )
  out$effects     <- eff_list
  out$num_effects <- vapply(eff_list, length, integer(1))
  out
})

effect_df_listcol <- do.call(rbind, effects_rows)

# ---- OPTIONAL: also build a wide padded matrix (effect_1..K) ----
make_wide <- TRUE
if (make_wide) {
  max_len <- if (nrow(effect_df_listcol)) max(effect_df_listcol$num_effects) else 0L
  if (max_len > 0) {
    effect_mat <- t(vapply(effect_df_listcol$effects, function(v) {
      len <- length(v)
      if (len < max_len) c(v, rep(NA_real_, max_len - len)) else v[seq_len(max_len)]
    }, numeric(max_len)))
    colnames(effect_mat) <- paste0("effect_", seq_len(ncol(effect_mat)))
    effect_df_wide <- cbind(effect_df_listcol[ , !(names(effect_df_listcol) %in% c("effects")), drop=FALSE],
                            as.data.frame(effect_mat, check.names = FALSE))
  } else {
    effect_df_wide <- effect_df_listcol[ , !(names(effect_df_listcol) %in% c("effects")), drop=FALSE]
  }
}

# ---- SAVE ----
saveRDS(effect_df_listcol, file = file.path(base_data_folder, paste0("ind/effect_df_gen", gen_filter, "_listcol.rds")))
if (make_wide) {
  saveRDS(effect_df_wide,    file = file.path(base_data_folder, paste0("ind/effect_df_gen", gen_filter, "_wide.rds")))
}

```

# load IND data
```{r}
# Load the cleaned effect data
effect_df_wide <- readRDS(
  file.path(base_data_folder, paste0("ind/effect_df_gen", gen_filter, "_wide.rds"))
)
```

## plot EFFECT SIZE histogram 
```{r}
# ---------------------------------------------------------
# EFFECT DISTRIBUTIONS BY PSI (from effect_df_wide)
# One PNG per (n, shift_size); averaged over individuals
# Base R only
# ---------------------------------------------------------
plot_ind_effect_dists_by_psi <- function(effect_df_wide,
                                         num_bins = 100,
                                         bin_breaks = NULL,   # if NULL, compute from the group's pooled values
                                         out_subfolder = "ind_effect_dists_by_psi",
                                         regimes = regimes) {

  if (!("reg" %in% names(effect_df_wide))) {
    stop("effect_df_wide must include 'reg' (e.g., 'ave' or 'fit').")
  }

  # effect columns
  xcols <- grep("^effect_", names(effect_df_wide), value = TRUE)
  if (!length(xcols)) stop("No effect_* columns found in effect_df_wide.")

  # helper: average normalized histogram over individuals for a given subset (single psi)
  avg_curve_for_subset <- function(df_block, breaks, num_bins) {
    if (nrow(df_block) == 0) return(rep(NA_real_, num_bins))
    counts_mat <- matrix(NA_real_, nrow = nrow(df_block), ncol = num_bins)
    for (i in seq_len(nrow(df_block))) {
      vals <- suppressWarnings(as.numeric(unlist(df_block[i, xcols])))
      vals <- vals[is.finite(vals)]
      if (!length(vals)) next
      h <- hist(vals, breaks = breaks, plot = FALSE)
      if (sum(h$counts) == 0) next
      counts_mat[i, ] <- h$counts / sum(h$counts)  # normalize per individual
    }
    colMeans(counts_mat, na.rm = TRUE)
  }

  # ---- loop by regime ----
  for (reg_lab in regimes) {
    df_reg <- subset(effect_df_wide, reg == reg_lab)
    if (nrow(df_reg) == 0) next

    # where to save (per regime)
    out_dir <- file.path(base_plot_folder, out_subfolder, paste0("reg_", reg_lab))
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

    for (n_val in sort(unique(df_reg$n))) {
      df_n <- subset(df_reg, n == n_val)
      if (nrow(df_n) == 0) next

      for (sh in sort(unique(df_n$shift_size))) {
        df_group <- subset(df_n, shift_size == sh)
        if (nrow(df_group) == 0) next

        # all psi values in this (reg, n, shift) group
        psi_vals <- sort(unique(df_group$psi))
        cols <- hcl.colors(length(psi_vals), palette = "Zissou 1")
        names(cols) <- as.character(psi_vals)

        # Choose bin breaks for THIS plot:
        #   use provided bin_breaks; otherwise compute from pooled effects across all individuals here
        if (is.null(bin_breaks)) {
          pooled <- suppressWarnings(as.numeric(unlist(df_group[, xcols])))
          pooled <- pooled[is.finite(pooled)]
          if (!length(pooled)) next
          rng <- range(pooled)
          if (!is.finite(rng[1]) || !is.finite(rng[2]) || rng[1] == rng[2]) {
            rng <- c(rng[1] - 0.5, rng[2] + 0.5)
          }
          breaks <- seq(rng[1], rng[2], length.out = num_bins + 1)
        } else {
          breaks <- bin_breaks
          # (optional safety) enforce expected length
          if (length(breaks) != num_bins + 1L) {
            warning("Provided bin_breaks length != num_bins + 1; recalculating from range.")
            pooled <- suppressWarnings(as.numeric(unlist(df_group[, xcols])))
            pooled <- pooled[is.finite(pooled)]
            if (!length(pooled)) next
            rng <- range(pooled)
            breaks <- seq(rng[1], rng[2], length.out = num_bins + 1)
          }
        }
        mids <- 0.5 * (breaks[-length(breaks)] + breaks[-1])

        # find a common y-limit across psi averages
        max_y <- 0
        for (psi_val in psi_vals) {
          avg_counts <- avg_curve_for_subset(subset(df_group, psi == psi_val), breaks, num_bins)
          max_y <- max(max_y, avg_counts, na.rm = TRUE)
        }
        if (!is.finite(max_y) || max_y <= 0) max_y <- 1

        # nicer label for shift_size
        title_shift <- if (is.finite(sh) && sh == 0) "data-driven" else paste0("shift_size = ", sh)
        file_shift  <- if (is.finite(sh) && sh == 0) "dataDriven"  else paste0("shift_", sh)

        # open PNG
        fname <- paste0("ind_effect_dists_reg_", reg_lab, "_n_", n_val, "_", file_shift, ".png")
        png(file.path(out_dir, fname), width = 1000, height = 700)
        par(mar = c(4.5, 4.5, 3.5, 1) + 0.1)

        # base empty plot
        plot(NULL,
             xlim = range(breaks), ylim = c(0, max_y),
             xlab = "Effect size", ylab = "Proportion",
             main = paste0("Effect distributions | reg = ", reg_lab,
                           " | n = ", n_val, " | ", title_shift))

        # draw one averaged curve per psi
        for (psi_val in psi_vals) {
          avg_counts <- avg_curve_for_subset(subset(df_group, psi == psi_val), breaks, num_bins)
          if (all(is.na(avg_counts))) next
          lines(mids, avg_counts, col = cols[as.character(psi_val)], lwd = 2)
        }

        legend("topright",
               legend = paste0("psi=", psi_vals),
               col = cols[as.character(psi_vals)],
               lty = 1, lwd = 2, cex = 0.9, bg = "white")

        dev.off()
      }
    }
  }
}

# ---- Run it ----
plot_ind_effect_dists_by_psi(
  effect_df_wide,
  num_bins = 40,
  bin_breaks = seq(-4, 4, by = 0.2),  # 8 / 40 = 0.2
  out_subfolder = "ind_effect_dists_by_psi",
  regimes = regimes  # e.g., c("ave","fit")
)
```


# load MUT data
```{r}
# ---------------------------------------------
# MUT files → per-generation mutation counts
# - no headers
# - variable-length rows/columns
# - odd rows: mutation IDs ... <gen>
# - even rows: effect sizes (ignored)
# ---------------------------------------------

mut_folder <- file.path(base_data_folder, "mut")

mut_files <- list.files(
  path = mut_folder,
  pattern = "\\.tsv$",
  full.names = TRUE,
  recursive = TRUE
)

# Filename pattern (UPDATED):
#   mut/n_{n}_psi_{psi}_sz_{shift}_reg_{ave|fit}_{rep}.tsv
extract_mut_metadata <- function(path) {
  rx <- "mut/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)\\.tsv"
  m  <- regexec(rx, path)
  p  <- regmatches(path, m)[[1]]
  if (length(p) != 6) {
    warning("Could not extract metadata from path: ", path, " | parts: ", paste(p, collapse=", "))
    return(NULL)
  }
  list(
    n          = as.integer(p[2]),
    psi        = as.numeric(p[3]),
    shift_size = as.integer(p[4]),
    reg        = p[5],                 # "ave" or "fit"
    rep        = as.integer(p[6]),
    file_id    = basename(path)
  )
}

# Utility: find index of the last non-empty, purely-integer token
.last_integer_token_idx <- function(tokens) {
  if (!length(tokens)) return(NA_integer_)
  nz <- which(nzchar(tokens))
  if (!length(nz)) return(NA_integer_)
  toks <- tokens[nz]
  for (k in seq_along(toks)) {
    idx <- length(toks) - k + 1L
    if (grepl("^-?\\d+$", toks[idx])) {
      return(nz[idx])  # map back to original position
    }
  }
  NA_integer_
}

count_mutations_in_file <- function(file) {
  lines <- readLines(file, warn = FALSE)
  if (!length(lines)) return(NULL)

  meta <- extract_mut_metadata(file)
  if (is.null(meta)) return(NULL)

  out <- vector("list", length = ceiling(length(lines) / 2))
  oi <- 0L

  # process odd rows only: 1,3,5,...
  for (i in seq.int(1L, length(lines), by = 2L)) {
    raw <- lines[i]
    if (!nzchar(raw)) next

    toks <- strsplit(raw, "\t", fixed = TRUE)[[1]]
    toks <- trimws(toks, which = "both")

    gen_pos <- .last_integer_token_idx(toks)
    if (is.na(gen_pos)) {
      warning("No numeric generation token found (odd row ", i, ") in: ", file)
      next
    }
    gen <- suppressWarnings(as.integer(toks[gen_pos]))
    if (!is.finite(gen)) next

    # Everything before gen_pos are mutation IDs (variable count, may include blanks)
    id_tokens <- if (gen_pos > 1L) toks[1:(gen_pos - 1L)] else character(0)
    mut_count <- sum(nzchar(id_tokens))

    oi <- oi + 1L
    out[[oi]] <- data.frame(
      n          = meta$n,
      psi        = meta$psi,
      shift_size = meta$shift_size,
      reg        = meta$reg,
      rep        = meta$rep,
      file_id    = meta$file_id,
      generation = gen,
      mut_count  = mut_count,
      stringsAsFactors = FALSE
    )
  }

  if (oi == 0L) return(NULL)
  do.call(rbind, out[seq_len(oi)])
}

# Run over all files
mut_count_df <- do.call(rbind, lapply(mut_files, count_mutations_in_file))

# Optional tidy-up
if (!is.null(mut_count_df) && nrow(mut_count_df)) {
  mut_count_df <- mut_count_df[order(mut_count_df$n,
                                     mut_count_df$psi,
                                     mut_count_df$reg,
                                     mut_count_df$shift_size,
                                     mut_count_df$rep,
                                     mut_count_df$generation), ]
}

```

## plot number of alleles over psi and n
```{r}
# ------------------------------------------------------
# Mutations vs psi (gen = 10000), one PNG per (n, shift_size)
# Base R only
# ------------------------------------------------------
plot_mut_counts_over_psi <- function(mut_count_df,
                                     gen_target = 10000L,
                                     out_subfolder = "mut_counts_by_psi",
                                     regimes = regimes) {
  if (!("reg" %in% names(mut_count_df))) {
    stop("mut_count_df must include 'reg' (e.g., 'ave' or 'fit').")
  }

  rep_col  <- adjustcolor("steelblue", alpha.f = 0.35)
  mean_col <- "steelblue4"

  for (reg_lab in regimes) {
    d_reg <- subset(mut_count_df, reg == reg_lab & generation == gen_target)
    if (nrow(d_reg) == 0) {
      warning("No rows at generation ", gen_target, " for reg=", reg_lab, " in mut_count_df.")
      next
    }

    out_dir <- file.path(base_plot_folder, out_subfolder, paste0("reg_", reg_lab))
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

    for (n_val in sort(unique(d_reg$n))) {
      d_n <- subset(d_reg, n == n_val)
      if (nrow(d_n) == 0) next

      for (sh in sort(unique(d_n$shift_size))) {
        d <- subset(d_n, shift_size == sh)
        if (nrow(d) == 0) next

        psi_grid <- sort(unique(d$psi))
        reps     <- sort(unique(d$rep))

        rep_series <- lapply(reps, function(r) {
          yv <- rep(NA_real_, length(psi_grid))
          rows <- d[d$rep == r, c("psi","mut_count")]
          if (nrow(rows)) {
            rows$psi <- as.numeric(rows$psi)
            agg <- aggregate(mut_count ~ psi, data = rows, FUN = mean)
            idx <- match(agg$psi, psi_grid)
            yv[idx] <- agg$mut_count
          }
          yv
        })
        names(rep_series) <- as.character(reps)

        if (length(rep_series)) {
          mat <- do.call(rbind, rep_series)
          mean_series <- apply(mat, 2L, function(col) mean(col, na.rm = TRUE))
        } else {
          mean_series <- rep(NA_real_, length(psi_grid))
        }

        y_all <- c(unlist(rep_series, use.names = FALSE), mean_series)
        y_all <- y_all[is.finite(y_all)]
        if (!length(y_all)) next
        y_min <- min(y_all); y_max <- max(y_all)
        if (y_min == y_max) { y_min <- y_min - 0.5; y_max <- y_max + 0.5 }

        # nicer label for shift_size
        title_shift <- if (is.finite(sh) && sh == 0) "data-driven" else paste0("shift_size = ", sh)
        file_shift  <- if (is.finite(sh) && sh == 0) "dataDriven"  else paste0("shift_", sh)

        fname <- paste0("mut_count_over_psi_reg_", reg_lab,
                        "_n_", n_val, "_", file_shift, "_gen", gen_target, ".png")
        png(file.path(out_dir, fname), width = 1000, height = 700)
        par(mar = c(4.5, 4.5, 3.5, 1) + 0.1)

        plot(NA,
             xlim = range(psi_grid), ylim = c(y_min, y_max),
             xlab = expression(psi), ylab = "Mutation count",
             main = paste0("Mutations vs psi | reg = ", reg_lab,
                           " | n = ", n_val, " | ", title_shift,
                           " | gen = ", gen_target),
             xaxt = "n")
        axis(1, at = psi_grid, labels = formatC(psi_grid, format = "f", digits = 1))

        if (length(rep_series)) {
          for (i in seq_along(rep_series)) {
            yv <- rep_series[[i]]
            lines(psi_grid, yv, col = rep_col, lwd = 1)
            points(psi_grid, yv, pch = 16, cex = 0.7, col = rep_col)
          }
        }

        lines(psi_grid, mean_series, col = mean_col, lwd = 2.5)
        points(psi_grid, mean_series, pch = 16, cex = 1, col = mean_col)

        legend("topleft", inset = 0.02,
               legend = c("Replicate", "Across-rep mean"),
               col = c(rep_col, mean_col), lty = c(1,1),
               lwd = c(1, 2.5), pch = c(16,16),
               pt.cex = c(0.7, 1.0), bty = "n", cex = 0.9)

        dev.off()
      }
    }
  }
}

# ---- Run ----
plot_mut_counts_over_psi(
  mut_count_df,
  gen_target = 10000L,
  out_subfolder = "mut_counts_by_psi",
  regimes = regimes  # e.g., c("ave","fit")
)

```

## plot population level mut distribution
```{r}
# --- Config (bins) ---
num_bins   <- 40
bin_breaks <- seq(-4, 4, length.out = num_bins + 1)
bin_mids   <- 0.5 * (bin_breaks[-1] + bin_breaks[-length(bin_breaks)])

# --- Files ---
mut_folder <- file.path(base_data_folder, "mut")
mut_files <- list.files(mut_folder, pattern = "\\.tsv$", full.names = TRUE, recursive = TRUE)

# --- Filename metadata (UPDATED):
# mut/n_{n}_psi_{psi}_sz_{shift}_reg_{ave|fit}_{rep}.tsv
extract_mut_metadata <- function(path) {
  rx <- "mut/n_(\\d+)_psi_(-?[0-9.]+)_sz_(\\d+)_reg_(ave|fit)_(\\d+)\\.tsv"
  m  <- regexec(rx, path)
  p  <- regmatches(path, m)[[1]]
  if (length(p) != 6) return(NULL)
  list(
    n          = as.integer(p[2]),
    psi        = as.numeric(p[3]),
    shift_size = as.integer(p[4]),
    reg        = p[5],                 # "ave" or "fit"
    rep        = as.integer(p[6]),
    file_id    = basename(path)
  )
}

# --- Helpers to parse lines with variable length & no headers ---
.last_integer_token_idx <- function(tokens) {
  if (!length(tokens)) return(NA_integer_)
  nz <- which(nzchar(tokens))
  if (!length(nz)) return(NA_integer_)
  toks <- tokens[nz]
  for (k in seq_along(toks)) {
    idx <- length(toks) - k + 1L
    if (grepl("^-?\\d+$", toks[idx])) return(nz[idx])
  }
  NA_integer_
}

# --- Build per-rep LONG table of normalized histograms ---
# One row per (file × generation × bin)
# Columns: n, psi, shift_size, reg, rep, file_id, generation, bin_mid, prop
mut_hist_long <- (function() {
  out <- vector("list", length(mut_files))
  oi  <- 0L

  for (file in mut_files) {
    meta <- extract_mut_metadata(file)
    if (is.null(meta)) next

    lines <- readLines(file, warn = FALSE)
    if (!length(lines)) next

    # Walk odd/even pairs
    for (i in seq.int(1L, length(lines), by = 2L)) {
      # ---- odd row: IDs ----
      ids_row <- lines[i]
      if (!nzchar(ids_row)) next
      ids_toks <- strsplit(ids_row, "\t", fixed = TRUE)[[1]]
      ids_toks <- trimws(ids_toks)
      gen_pos1 <- .last_integer_token_idx(ids_toks)
      if (is.na(gen_pos1)) next
      gen1 <- suppressWarnings(as.integer(ids_toks[gen_pos1]))
      if (!is.finite(gen1)) next

      # ---- even row: effects ----
      if (i + 1L > length(lines)) next
      eff_row <- lines[i + 1L]
      eff_toks <- strsplit(eff_row, "\t", fixed = TRUE)[[1]]
      eff_toks <- trimws(eff_toks)
      gen_pos2 <- .last_integer_token_idx(eff_toks)
      gen2 <- if (!is.na(gen_pos2)) suppressWarnings(as.integer(eff_toks[gen_pos2])) else NA_integer_
      # prefer even-row gen if valid; else use odd-row gen
      gen <- if (is.finite(gen2)) gen2 else gen1
      if (!is.finite(gen)) next

      # effect tokens are everything before the gen token on the even row
      eff_tokens <- if (!is.na(gen_pos2) && gen_pos2 > 1L) eff_toks[1:(gen_pos2 - 1L)] else character(0)
      # fallback: if even row malformed, try IDs row length
      if (!length(eff_tokens) && gen_pos1 > 1L) eff_tokens <- ids_toks[1:(gen_pos1 - 1L)]

      eff_vals <- suppressWarnings(as.numeric(eff_tokens))
      eff_vals <- eff_vals[is.finite(eff_vals)]
      if (!length(eff_vals)) next  # no effects → skip this generation for this rep

      h <- hist(eff_vals, breaks = bin_breaks, plot = FALSE)
      if (sum(h$counts) == 0) next
      prop <- h$counts / sum(h$counts)

      # add long rows (one per bin)
      df_piece <- data.frame(
        n          = meta$n,
        psi        = meta$psi,
        shift_size = meta$shift_size,
        reg        = meta$reg,
        rep        = meta$rep,
        file_id    = meta$file_id,
        generation = gen,
        bin_mid    = bin_mids,
        prop       = prop,
        stringsAsFactors = FALSE
      )

      oi <- oi + 1L
      out[[oi]] <- df_piece
    }
  }
  if (oi == 0L) return(data.frame())
  do.call(rbind, out[seq_len(oi)])
})()

# --- Average across reps (mean proportion per bin) ---
# Gives one row per (n, psi, shift_size, reg, generation, bin_mid)
if (nrow(mut_hist_long)) {
  mut_hist_mean <- aggregate(
    prop ~ n + psi + shift_size + reg + generation + bin_mid,
    data = mut_hist_long,
    FUN = function(x) mean(x, na.rm = TRUE)
  )
  # (optional) how many reps contributed to each mean
  mut_hist_n <- aggregate(
    prop ~ n + psi + shift_size + reg + generation + bin_mid,
    data = mut_hist_long,
    FUN = function(x) sum(is.finite(x))
  )
  names(mut_hist_n)[names(mut_hist_n) == "prop"] <- "n_reps"
  mut_hist_mean <- merge(
    mut_hist_mean, mut_hist_n,
    by = c("n","psi","shift_size","reg","generation","bin_mid"),
    all.x = TRUE, sort = FALSE
  )
} else {
  mut_hist_mean <- data.frame()
}

```

```{r}
# ---------------------------------------------
# Plot averaged mut effect-size hists for gen 10000 and 10005
# One PNG per (n, shift_size, generation); lines = psi
# ---------------------------------------------
plot_mut_hist_two_gens <- function(mut_hist_mean,
                                   gens = c(10000, 10005),
                                   out_subfolder = "mut_effect_hist_avg_two_gens",
                                   regimes = regimes) {
  if (!nrow(mut_hist_mean)) {
    warning("mut_hist_mean is empty.")
    return(invisible(NULL))
  }
  if (!("reg" %in% names(mut_hist_mean))) {
    stop("mut_hist_mean must include 'reg' (e.g., 'ave' or 'fit').")
  }

  for (reg_lab in regimes) {
    df_reg <- subset(mut_hist_mean, reg == reg_lab)
    if (nrow(df_reg) == 0) next

    out_dir <- file.path(base_plot_folder, out_subfolder, paste0("reg_", reg_lab))
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

    for (n_val in sort(unique(df_reg$n))) {
      df_n <- subset(df_reg, n == n_val)
      if (nrow(df_n) == 0) next

      for (sh in sort(unique(df_n$shift_size))) {
        df_ns <- subset(df_n, shift_size == sh)
        if (nrow(df_ns) == 0) next

        # Nice label/filename piece for shift_size
        title_shift <- if (is.finite(sh) && sh == 0) "data-driven" else paste0("shift_size = ", sh)
        file_shift  <- if (is.finite(sh) && sh == 0) "dataDriven"  else paste0("shift_", sh)

        for (g in gens) {
          d <- subset(df_ns, generation == g)
          if (nrow(d) == 0) next

          psi_vals <- sort(unique(d$psi))
          cols <- hcl.colors(length(psi_vals), palette = "Zissou 1")
          names(cols) <- as.character(psi_vals)

          # y-limit from data
          y_max <- max(d$prop, na.rm = TRUE)
          if (!is.finite(y_max) || y_max <= 0) y_max <- 1

          # x-limits from bin mids (expand slightly if we can estimate bin width)
          mids_all <- sort(unique(d$bin_mid))
          if (length(mids_all) > 1) {
            bw <- min(diff(mids_all))
            xlim <- c(min(mids_all) - 0.5 * bw, max(mids_all) + 0.5 * bw)
          } else {
            xlim <- range(mids_all)
          }

          # File + device
          fname <- paste0("mut_avg_hist_reg_", reg_lab, "_n_", n_val, "_", file_shift, "_gen", g, ".png")
          png(file.path(out_dir, fname), width = 1000, height = 700)
          par(mar = c(4.5, 4.5, 3.5, 1) + 0.1)

          plot(NULL, xlim = xlim, ylim = c(0, y_max),
               xlab = "Effect size", ylab = "Proportion",
               main = paste0("Unique-mutation effects (avg over reps)\n",
                             "reg = ", reg_lab, " | n = ", n_val, " | ",
                             title_shift, " | gen = ", g))

          # Draw one averaged curve per psi
          for (psi_val in psi_vals) {
            dd <- d[d$psi == psi_val, ]
            dd <- dd[order(dd$bin_mid), ]
            if (!nrow(dd)) next
            lines(dd$bin_mid, dd$prop, col = cols[as.character(psi_val)], lwd = 2)
          }

          legend("topright",
                 legend = paste0("psi=", psi_vals),
                 col = cols[as.character(psi_vals)],
                 lty = 1, lwd = 2, cex = 0.9, bg = "white")

          dev.off()
        }
      }
    }
  }
}

# Run
plot_mut_hist_two_gens(
  mut_hist_mean,
  gens = c(10000, 10005),
  out_subfolder = "mut_effect_hist_avg_two_gens",
  regimes = regimes  # e.g., c("ave","fit")
)

```
