---
title: "plot genotype data for multiple models"
author: "Andrea Chen"
date: "`r Sys.Date()`"
output: pdf_document
---

# parameters for plotting
```{r setup, include=FALSE}
wd_path <- normalizePath(dirname(rstudioapi::getActiveDocumentContext()$path))
setwd(wd_path)
base_plot_folder <- "../plots/evolve_sociality"
hcl_color_palette <- "Dark 2" # color pattern for lines 
param_val <- c(0.0) # for now no base psi and beta
```

# load genotype data
```{r}
psis_betas_df <- readRDS("../r_data/evolve_sociality_sociality.rds")
```

# parameter space for data
regimes <- c("inv", "fit", "ext", "ave")
## all group sizes
plot_n <- c(2, 8, 20, 50)
## all shift sizes 
plot_shift_sizes <- c(40,80,12, 300) 
## all generations
gen_seq = c(seq(9999, 10010, by = 1), seq(10010, 10100, by=5)) # generations to plot

# change parameters here for plotting a subset of the data
## PARAMETER: single panel single generation plot

```{r setup, include=FALSE}
param_val <- c(0.0) 
plot_n <- c(2, 8, 20, 50) # group sizes to plot for both models
plot_shift_sizes <- c(40,80,120, 300) # shift sizes to plot for both models
regimes <- c("ave")  # "inv" = inverse variance model, "fit" = fittest individual model
single_panel <- TRUE  
num_bins <- 80
gen_seq = seq(9999, 9999, by = 1)
```

# RUN PLOTTING after change the parameters above
```{r}
min_bin <- -as.integer(num_bins/2)
max_bin <- as.integer(num_bins/2)
bin_breaks <- seq(min_bin, max_bin, length.out = num_bins + 1)

# Derive a constant bin width from your reference breaks
bin_width <- (max(bin_breaks) - min(bin_breaks)) / num_bins
half_span <- (num_bins * bin_width) / 2

if (regimes != "inv") {
  plot_psi <- param_val
} else {
  plot_beta <- param_val
}

# ---- Run ----
# Example call:
plot_psi_histograms_by_gen(
  df = psis_betas_df,
  target_gen = 10050,
  regs = c("ext","ave","fit"),
  plot_n = c(2,8, 20, 50),         # or NULL to auto-detect per reg
  pool_reps = TRUE,         # TRUE = pool reps; FALSE = bin-per-rep then average
  value_col = "value",              # change if your ψ column has a different name
  nbins = 100,                   # or provide bin_breaks = seq(-1,1,by=0.02)
  out_subfolder = "psi_dist_onegen"
)

```


```{r}
min(psis_betas_df[psis_betas_df$reg == "fit", ]$value)

```